# rebi - R Interface to Europe PMC RESTful Web Service

```{r echo=FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  warning = FALSE,
  message = FALSE
)
```

[![Build Status](https://travis-ci.org/njahn82/rebi.svg?branch=master)](https://travis-ci.org/njahn82/rebi)

rebi facilitates access to [Europe PMC RESTful Web Service](http://europepmc.org/RestfulWebService). 

Europe PubMed Central, one of the largest literature databases for life-science 
publications,  indexes more than 30 millions publications, including 25
millions from PubMed. It gives also access to full texts, if available.

## Installation

The latest development version can be installed using [devtools](https://github.com/hadley/devtools) package:

```{r, eval = FALSE}
require(devtools)
install_github("njahn82/rebi")
```

## Search Europe PMC

Europe PMC provides comprehensive guidance on how to search:
<http://europepmc.org/help>

```{r}
my_data <- epmc_search(query='Gabi-Kat')
# number of records
my_data$hit_count
# first six records
head(my_data$md)
```

Get PLOS Genetics articles that cross-reference EMBL

```{r}
my_data <- epmc_search(query = 'ISSN:1553-7404 HAS_EMBL:y')
head(my_data$md)
```

Get list of ids that represent articles referencing DataCite DOIs

```{r}
my_data <- epmc_search(query = "ACCESSION_TYPE:doi", n_pages = 1, 
                       id_list = TRUE)
head(my_data$md)
my_data$hit_count
```

Search by ORCID

```{r}
my_data <- epmc_search(query = 'AUTHORID:"0000-0002-7635-3473"', n_pages = 1) 
head(my_data$md)
my_data$hit_count
```

## Get article details

```{r}
epmc_details(ext_id = "24270414")
```

Show author details including orcid, if available

```{r}
my_data <- epmc_details(ext_id = "14756321")
my_data$author_details
```

## Get citation counts and citing publications

```{r}
my_cites <- epmc_citations("9338777")
my_cites$hit_count
head(my_cites$citations)
```

## Get reference section

```{r}
epmc_refs("PMC3166943", data_src = "pmc")
```

## Retrieve links to other EBI databases

Summary of links found

```{r}
epmc_db_count("12368864")
```

Select database and get links:

```{r}
epmc_db("12368864", db = "embl")
```


## Get text-mined terms

Summary of text-mined terms

```{r}
epmc_tm_count("25249410")
```

Select semantic type to retrieve the terms:

```{r}
epmc_tm("25249410", semantic_type = "GO_TERM")
```


## References

Europe PMC: a full-text literature database for the life sciences and platform 
for innovation. (2014). Nucleic Acids Research, 43(D1), D1042–D1048. doi:[10.1093/nar/gku1061](http://doi.org/10.1093/nar/gku1061)
