---
title: "#evergreenreview graphs"
author: "Najko Jahn"
date: "15 May 2017"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Overview}
  \usepackage[utf8]{inputenc}
---

```{r, echo=FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = TRUE,
  fig.width = 12,
  fig.height = 8
)
options(scipen = 999, digits = 2)
knitr::knit_hooks$set(inline = function(x) {
      if(is.numeric(x)){
          return(prettyNum(x, big.mark=","))
      }else{
          return(x)
       }
   })
```

## What is a evergreen review graph?

<http://www.masalmon.eu/2017/05/14/evergreenreviewgraph/>

## Making proper review graphs with Europe PMC

### Growth of Open Access Literature

There is growing interest in knowing the proportion of open access to scholarly literature. Europe PMC allows searching for open access content with with the [OPEN_ACCESS:Y parameter](https://europepmc.org/search?query=OPEN_ACCESS:Y&page=1&sortby=Relevance). At the moment, Europe PMC contains `r europepmc::epmc_hits("OPEN_ACCESS:Y")` open access fulltexts. Let's see how they are relatively distributed over the period 1995 - 2016.

```{r, fig.align='center', fig.width=7, fig.height=4}
library(ggplot2)
tt_oa <- europepmc::epmc_hits_trend("OPEN_ACCESS:Y", period = 1995:2016, synonym = FALSE)
ggplot(tt_oa, aes(year, query_hits / all_hits)) + 
  geom_point() + 
  geom_line() +
  xlab("Year published") + 
  ylab("Proportion of OA full-texts in Europe PMC")
```

Be careful with your interpretion of the last years as Europe PMC also contains journals, which may make their articles openly availabe after an embargo period (up to two years). 
